Evaluating impacts of Density Dependent processes on Blue whiting via MSE simulations

### 2023-03-17 Rmd/report mac.Rmd
Fixed error in figure 15

Version 1.3 (`r format(Sys.time(), '%d %B %Y')`)

@@

**Comparison of equilibrium curves for density dependence analyses: biomass against yield and F**

The comparison of the equilibrium curves, indicate that B_MSY is expected to decrease by 30%, $F_{MSY}$ is expected to increase by 113% and catch (MSY) is expected to increase by 5%. So while, inclusion of density-dependent processes, has a substantial impact on biological reference points, the resulting change in equilibrium catch is expected to be small.

@@

**Reference points**

Reference points estimated for the different scenarios are in the table below.

Forward projections were run in two modes. Firstly in deterministic mode and secondly in probabilistic mode. We used the $F_{MSY}$ to project forward from 2020 until 2050. We used both the $F_{MSY}$ that is applicable to the density dependent scenario, but also evaluated what would happen if the 'wrong $F_{MSY}$' would be used with a particular scenario.

>>>>

Version 1.3 (13 March 2023)

Version 1.4 (16 March 2023; fixed error in reference point table)

Version 1.5 (17 March 2023; fixed error in figure 15)

@@

Forward projections were run in two modes. Firstly in deterministic mode and secondly in probabilistic mode. We used the $F_{MSY}$ to project forward from 2020 until 2050. The reference points estimated for the different scenarios are in the table below.

The comparison of the equilibrium curves, indicate that B_MSY is expected to decrease by 35%, $F_{MSY}$ is expected to increase by factor of 2.5 and catch (MSY) is expected to increase by 9%. So while, inclusion of density-dependent processes, has a substantial impact on biological reference points, the resulting change in equilibrium catch is expected to be small.

@@

We used both the $F_{MSY}$ that is applicable to the density dependent scenario, but also evaluated what would happen if the 'wrong $F_{MSY}$' would be used with a particular scenario.

### 2023-03-17 Rmd/report whb.Rmd
Fixed error in figure 15

Version 1.0 (`r format(Sys.time(), '%d %B %Y')`)

@@

**Laurie: why would DD in natural mortality have such a minimal impact? Did you see that the M levels overall are lower than the assumed M in the assessment (=0.2)**

@@

Initial evaluations were conducted based on life history theory, as used by the International Council for the Exploration of the Sea (ICES) to develop advice conducting MSE for data poor stocks [@fischer2020ga]. **Figure 10** shows the the equilibrium curves of catch v SSB assuming a) no density dependence, and density dependence in b) mass-at-age, c) mass & maturity-at-age, and d) mass, maturity and natural mortality-at-age. These allow the relationships between the MSY reference points and virgin biomass to be seen.

@@

The OMs were also projected for the estimates of different estimates $F_{MSY}$, with stochasticty in the recruitment deviates. This allows the impact of misspecification of density dependence to be evaluated, i.e. if $F_{MSY}$ is underestimated then yield will be less than MSY, while if $F_{MSY}$ is overestimated yield will be lost due to overfishing. Stochastic projections of the four Operating Models for the four estimates of $F_{MSY}$ are shown in **Figures 16-19**. **Figures 20-22** summarizes the yields, SSBs and Fs obtained for each Operating Model under the correct and mis-specified form of density dependence.

@@

Four alternatives OMs have been included in the analyses:

@@

All MSE's have been developed within the FLR framework (<https://flr-project.org/>) and all the code is stored on github (<https://github.com/martinpastoors/ddmse>).

@@

Mass-at-age ($W_a$) is modeled as

@@

Maturity is then modeled as a logistic function of mass-at-age i.e.

@@

@@

**Comparison of equilibrium curves for density dependence analyses: biomass against yield and F**

The comparison of the equilibrium curves, indicate that B_MSY is expected to decrease by 30%, $F_{MSY}$ is expected to increase by 113% and catch (MSY) is expected to increase by 5%. So while, inclusion of density-dependent processes, has a substantial impact on biological reference points, the resulting change in equilibrium catch is expected to be small.

@@

Misspecification of Fmsy, i.e. using an Fmsy from one scenario and applying it to a different scenario has only limited impacts on the overall outcomes in terms of recruitment, biomass or yield. In the plot below, the scenarios for density dependence are shown in the columns and the Fmsy's are shown as the colors. For example a valid combination is the Base case with the red curves (Fmsy=0.27) or the M scenario with the green curve. Overall the misspecification does not results in noticeable differences in catch although they do have important difference for biomass.

@@

Finally, as a check, we assessed the outcomes of the equilibrium curves from simulating over a wide F range, with the equilibrium values (points) in 2050 from the forward projections at different values of Fmsy (colors). The points fall on the equilibrium curve, thereby confirming the consistency between the two approaches.

@@

Initial explorations of density dependent effect were conducted based on life history theory, as used by the International Council for the Exploration of the Sea (ICES) to develop advice conducting MSE for data poor stocks [@fischer2020ga]. **Figure 10** shows the the equilibrium curves of catch v SSB assuming a) no density dependence, and density dependence in b) mass-at-age, c) mass & maturity-at-age, and d) mass, maturity and natural mortality-at-age. These allow the relationships between the MSY reference points and virgin biomass to be seen.

@@

In the next step, we applied a natural mortality-at-age vector based on [@lorenzen2000allometry] which mimics the VPA assessment with fixed M-at-age. The figure below compares the SAM assessment with a Virtual Population Analysis (VPA) and a VPA with age varying M (VPAM). Trends and absolute levels swere similar across all runs. The vpaM model was used as the base case to be compared against the density-dependent simulations.

@@

The OMs were also projected for the four estimates of  $F_{MSY}$, with stochasticty in the recruitment deviates. This allows the impact of misspecification of density dependence to be evaluated, i.e. if $F_{MSY}$ is underestimated then yield will be less than MSY, while if $F_{MSY}$ is overestimated yield will be lost due to overfishing. Stochastic projections of the four Operating Models for the four estimates of $F_{MSY}$ are shown in **Figures 16-19**

@@

The quality of this analyses lives up to the quality standards that are required by ICES and so can potentially be used when providing advice to fisheries managers. The software is publicly available via Github (<https://github.com/martinpastoors/ddmse>). All data is derived from WGWIDE 2022 @ICES2022WGWIDE.

>>>>

Version 1.0 (15 March 2023)

Version 1.1 (16 March 2023, fixed error in reference point table)

Version 1.2 (17 March 2023; fixed error in figure 15)

@@

@@

Initial evaluations were conducted based on life history theory, as used by the International Council for the Exploration of the Sea (ICES) to develop advice conducting MSE for data poor stocks [@fischer2020ga]. **Figure 10** shows the equilibrium curves of catch v SSB assuming a) no density dependence, and density dependence in b) mass-at-age, c) mass & maturity-at-age, and d) mass, maturity and natural mortality-at-age. These allow the relationships between the MSY reference points and virgin biomass to be seen.

@@

The OMs were also projected for the estimates of different estimates $F_{MSY}$, with stochasticity in the recruitment deviates. This allows the impact of misspecification of density dependence to be evaluated, i.e. if $F_{MSY}$ is underestimated then yield will be less than MSY, while if $F_{MSY}$ is overestimated yield will be lost due to overfishing. Stochastic projections of the four Operating Models for the four estimates of $F_{MSY}$ are shown in **Figures 16-19**. **Figures 20-22** summarizes the yields, SSBs and Fs obtained for each Operating Model under the correct and mis-specified form of density dependence.

@@

Four alternative OMs have been included in the analyses:

@@

All MSE's have been developed within the FLR framework (<https://flr-project.org/>) and all the code is stored on GitHub (<https://github.com/martinpastoors/ddmse>).

@@

Mass-at-age ($W_a$) is modelled as

@@

Maturity is then modelled as a logistic function of mass-at-age i.e.

@@

Forward projections were run in two modes. Firstly in deterministic mode and secondly in probabilistic mode. We used the $F_{MSY}$ to project forward from 2020 until 2050. Reference points estimated for the different scenarios are in the table below.

@@

The comparison of the equilibrium curves, indicate that B_MSY is expected to decrease by 45%, $F_{MSY}$ is expected to increase by a factor of 5 and catch (MSY) is expected to increase by 8%. So, while inclusion of density-dependent processes has a substantial impact on biological reference points, the resulting change in equilibrium catch is expected to be small.

@@

Misspecification of Fmsy, i.e. using an Fmsy from one scenario and applying it to a different scenario has only limited impacts on the overall outcomes in terms of recruitment, biomass or yield. In the plot below, the scenarios for density dependence are shown in the columns and the Fmsy's are shown as the colours. For example, a valid combination is the Base case with the red curves (Fmsy=0.27) or the M scenario with the green curve. Overall, the misspecification does not results in noticeable differences in catch although they do have important difference for biomass.

@@

Finally, as a check, we assessed the outcomes of the equilibrium curves from simulating over a wide F range, with the equilibrium values (points) in 2050 from the forward projections at different values of Fmsy (colours). The points fall on the equilibrium curve, thereby confirming the consistency between the two approaches.

@@

Initial explorations of density dependent effect were conducted based on life history theory, as used by the International Council for the Exploration of the Sea (ICES) to develop advice conducting MSE for data poor stocks [@fischer2020ga]. **Figure 10** shows the equilibrium curves of catch v SSB assuming a) no density dependence, and density dependence in b) mass-at-age, c) mass & maturity-at-age, and d) mass, maturity and natural mortality-at-age. These allow the relationships between the MSY reference points and virgin biomass to be seen.

@@

In the next step, we applied a natural mortality-at-age vector based on [@lorenzen2000allometry] which mimics the VPA assessment with fixed M-at-age. The figure below compares the SAM assessment with a Virtual Population Analysis (VPA) and a VPA with age varying M (VPAM). Trends and absolute levels were similar across all runs. The vpaM model was used as the base case to be compared against the density-dependent simulations.

@@

The OMs were also projected for the four estimates of  $F_{MSY}$, with stochasticity in the recruitment deviates. This allows the impact of misspecification of density dependence to be evaluated, i.e. if $F_{MSY}$ is underestimated then yield will be less than MSY, while if $F_{MSY}$ is overestimated yield will be lost due to overfishing. Stochastic projections of the four Operating Models for the four estimates of $F_{MSY}$ are shown in **Figures 16-19**

@@

The quality of this analyses lives up to the quality standards that are required by ICES and so can potentially be used when providing advice to fisheries managers. The software is publicly available via GitHub (<https://github.com/martinpastoors/ddmse>). All data is derived from WGWIDE 2022 @ICES2022WGWIDE.
